### 2013-07-18 Tex/chapters/introduction.tex
Theorieteil zu Monte Carlo hinzugefügt

The computing capabilities of modern computers enable researchers to
collect and analyze vast amounts of data. The prime example for is fact
is the LHC, where petabytes of particle collision data is collected and
processed. But computers are not only able to process measured data, but
also to generate data by simulations of e.g. statistical systems, where
the rules for each subsystem are well defined but the behavior of the
whole system is not easy to predict, beacause of the interactions of the
subsystems. For those systems there often exist no analytic solutions
or only for very simplified or special cases. But one can simulate all
the interactions of the subsystems and observe the behavior of the whole
system.
For example to examine phase transitions, which are defined by the abrupt
change of an observable e.g. the change of the density of water near
boiling or the change of the magnetization of a ferromagnet near the
Curie temperature. One of the simplest models with a second order
phase transition is the Ising model \cite{Ising1925} which is a simple
model of a ferromagnet and will be explained in more detail in section
\ref{ssec:isingmodel}. It is analytically solved for two dimensions on
some regular lattices \cite{Onsager1944} \cite{Wannier1945}.
In this thesis its behavior near the Curie temperature -- also called
critical temperature -- and the behavior of the Curie temperature itself
on some irregular lattices corresponding to proximity graphs
(see section \ref{ssec:graphtypes}) will be examined using the Monte
Carlo simulations described in section \ref{sec:montecarlo}.

>>>>

The computing capabilities of modern computers enable researchers to
collect and analyze vast amounts of data. The prime example for is fact
is the LHC, where petabytes of particle collision data is collected and
processed. But computers are not only able to process measured data, but
also to generate data by simulations of e.g. statistical systems, where
the rules for each subsystem are well defined but the behavior of the
whole system is not easy to predict, beacause of the interactions of the
subsystems. For those systems there often exist no analytic solutions
or only for very simplified or special cases. But one can simulate all
the interactions of the subsystems and observe the behavior of the whole
system.
For example to examine phase transitions, which are defined by the abrupt
change of an observable e.g. the change of the density of water near
boiling or the change of the magnetization of a ferromagnet near the
Curie temperature. One of the simplest models with a second order
phase transition is the Ising model \cite{Ising1925} which is a simple
model of a ferromagnet and will be explained in more detail in section
\ref{ssec:isingmodel}. It is analytically solved for two dimensions on
some regular lattices \cite{Onsager1944} \cite{Wannier1945}.
In this thesis its behavior near the Curie temperature -- also called
critical temperature -- and the behavior of the Curie temperature itself
on some irregular lattices corresponding to proximity graphs
(see section \ref{ssec:graphtypes}) will be examined using the Monte
Carlo simulations described in section \ref{ssec:montecarlo}.

### 2013-07-18 Tex/chapters/methods.tex
Theorieteil zu Monte Carlo hinzugefügt

\label{sec:montecarlo}
%~ Monte Carlo simulations are (probably) named after the casino in Monaco
%~ \cite{NewmanBarkema1999} because both rely on randomness.\\
The idea behind Monte Carlo simulations is to take random samples of
the observable, which should be measured, and estimate the observable from
this samples. In example the Monte Carlo integration chooses random
points within the bounding box of the function and estimates the integral
as the fraction of points below the function and points above multiplied
by the area of the bounding box.
Of course one wants to sample more points, where the area below the
function is big, because points in areas, where the function does
contribute little to the integral, contribute little to the integral.
This can be achieved by \emph{Importance Sampling}. One
generates random numbers according to some probability distribution, which
yields more numbers in the range where the function to integrate has high
values and weights the result accordingly.\\
%~ To improve this method: \emph{Importance Sampling}
%~ means that one takes more samples where the contribution of the
%~ function to the integral is bigger by using random numbers
%~ distributed........
The same principle is used to sample properties of statistical systems
by generating random states.
But it is difficult to create a random state of a physical system e.g. the
Ising system in equilibrium, hence one uses \emph{Markov Chains} to
generate new states \(\nu\) from old ones \(\mu\).
It is important that the transition probabilities \(A(\mu \to \nu)\)
obey \emph{Detailed Balance} and \emph{Ergodicity}.
\emph{Detailed Balance} means  that the probability to leave a state is
the same as the probability to enter the state in equilibrium
\(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
probability to be in state \(\mu\).
And \emph{Ergodicity} requires that every possible state is reachable
from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
Otherwise the samples might not be representative for the whole system.\\

\subsection{Monte Carlo Algorithms}
    \subsubsection{Metropolis}
        A Metropolis Monte Carlo\cite{Metropolis1953} simulation of an
        Ising model will choose a random spin, calculate the energy change
        \(\Delta \hat H\) defined in eq. \eqref{eq:dH} that would result
        from a flip of that spin and execute the flip with the probability \(A\)
        eq. \eqref{eq:metropolis} \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
        \begin{equation}
            \Delta \hat H = \hat H(\nu) - \hat H(\mu)\\
            \label{eq:dH}
        \end{equation}
        \begin{equation}
            A(\mu \to \nu)
            \begin{cases}
                1                                 & \Delta \hat H \le 0 \\
                \exp{\brac{-\beta \Delta \hat H}} & \Delta \hat H > 0
            \end{cases}
            \label{eq:metropolis}
        \end{equation}
        So if a transition lowers the energy it is always done. This
        results in a high ratio between chosen spins and flipped spins.
        Therefore it minimizes the calculations needed for a change of
        the state.

@@

\subsection{Equilibration- and Autocorrelation Time}
\label{ssec:eqtime}
    To generate an equilibrium state one starts with an arbitrary state
    and waits until it is equilibrated. The count of sweeps till
    equilibrium is called \emph{equilibration time} \(t_{eq}\).
    All measurement should start after this time.\\
    To determine when the system is in equilibrium, one can watch the
    development of some observables and take the point at which there
    are no big changes anymore as equilibrium. In fig.
    \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
    the equilibrium is reached after approximately \(N=50\) sweeps for
    either an initial condition of all spins up and all spins random. It
    does not harm to double that value to be save.
    \begin{figure}[htbp]
        \centering
        \subfigure[Example of an equilibrating Ising system][]{
                \label{sfig:equiandauto:equiE}
                \includegraphics[width=0.47\textwidth]{plots/equiE}
        }
        \subfigure[Example of the autocorrelation of an Ising system][]{
                \label{sfig:equiandauto:autoM}
                \includegraphics[width=0.47\textwidth]{plots/autoM}
        }
        \caption[Examples for equilibration and autocorrelation]
        {
            \subref{sfig:equiandauto:equiE} Example of a Ising system
                \(L=64\) reaching equilibrium at \(T=2.36\) and
            \subref{sfig:equiandauto:autoM} the autocorrelation of an
                Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                sweeps -- otherwise the decline is too steep to show)
                on half logarithmic axis.
                The straight line is an exponential fit \(\exp(-t/\tau)\)
                with \(\tau = 342(1)\).
        }
        \label{fig:equiandauto}
    \end{figure}\\
    Because every state is generated from the state before, measurements
    of subsequent states are correlated. To determine when two states
    are independent, one calculates the normalized autocorrelation function
    \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
    \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
    logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
    To get the autocorrelation time one can either fit a exponential
    function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
    or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
    \(\tau\) is an estimate after which time two samples are not
    correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
    To make sure that the error is not underestimated one should wait
    \(2\tau\) sweeps between two measurements.
    The autocorrelation time is of course dependent on the temperature.
    For example for the standard Metropolis algorithm the fluctuations
    are strong at high temperatures and subsequent
    states are more dissimilar and therefore less correlated than at low
    temperatures, where less spins are flipping. But the longest
    autocorrelation times are encountered at the critical temperature.
    This effect is called \emph{critical slowing down} and is
    characterized by the \emph{dynamical critical exponent} \(z\)
    \cite{SwendsenWang1987}. The dependence of the autocorrelation time
    \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).
    As mentioned in section \ref{sssec:wolff}, the Wolff cluster algorithm
    decreases \(z\) dramatically. This causes the course of the autocorrelation
    time in dependence on temperature to change significantly as shown in
    section \ref{sec:results}.

>>>>

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    To examine the disordered Ising model in Equilibrium,
    Its temperature \(T\),
    volume and number of particles \(N\) are constant, so that it is a canonical
    ensemble. In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \[p_i \propto e^{-\frac{\hat{H}}{k_B T}}\]
    Further the free energy \(F\) of an canonical
    ensemble is minimized and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g. \cite{nolting2005}).
    To determine \(F\), one has to know the partition function
    \[Z=\sum_i e^{-\frac{\hat{H}_i}{kT}}\]
    where the sum goes over all possible states \(i\) of the system.
    For small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But then there are of course next to
    infinity many different states, such that it is unfeasible to
    calculate the energy for each state, except for cases, where it is
    possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations.\\

\subsection{Monte Carlo Simulations}
\label{ssec:montecarlo}
    The idea behind Monte Carlo simulations is to take random samples of
    the observable, which should be measured, and estimate the observable from
    this samples. In example the Monte Carlo integration chooses random
    points within the bounding box of the function and estimates the integral
    as the fraction of points below the function and points above multiplied
    by the area of the bounding box.
    Of course one wants to sample more points, where the area below the
    function is big, because points in areas, where the function does
    contribute little to the integral, contribute little to the integral.
    This can be achieved by \emph{Importance Sampling}. One
    generates random numbers according to some probability distribution, which
    yields more numbers in the range where the function to integrate has high
    values and weights the result accordingly.\\
    The same principle is used to sample properties of statistical
    systems by generating random states.
    As noted before the system under scrutiny is canonical and therefore
    the states are distributed according to the Boltzman distribution.
    Hence importance sampling can be utilized.\\
    But it is difficult to create a random state of a physical system e.g. the
    Ising system in equilibrium, hence one uses \emph{Markov Chains} to
    generate new states \(\nu\) from old ones \(\mu\).
    It is important that the transition probabilities \(A(\mu \to \nu)\)
    obey \emph{Detailed Balance} and \emph{Ergodicity}.
    \emph{Detailed Balance} means that the probability to leave a state is
    the same as the probability to enter the state in equilibrium
    \(p_\mu A(\mu \to \nu) = p_\nu A(\nu \to \mu)\) with \(p_\mu\) the
    probability to be in state \(\mu\).
    And \emph{Ergodicity} requires that every possible state is reachable
    from every other state in finite time. \cite{NewmanBarkema1999} \cite{Katzgraber2011}
    Otherwise the samples might not be representative for the whole system.\\
    In this thesis three algorithms, which fulfill all requirements,
    were used. They will be shortly described in the following subsections.

    \subsubsection{Metropolis}
        A Metropolis Monte Carlo\cite{Metropolis1953} simulation of an
        Ising model will choose a random spin, calculate the energy change
        \(\Delta \hat H\) defined in eq. \eqref{eq:dH} that would result
        from a flip of that spin and execute the flip with the probability \(A\)
        eq. \eqref{eq:metropolis} \cite{NewmanBarkema1999} \cite{Katzgraber2011}.
        \begin{equation}
            \Delta \hat H = \hat H(\nu) - \hat H(\mu)\\
            \label{eq:dH}
        \end{equation}
        \begin{equation}
            A(\mu \to \nu)
            \begin{cases}
                1                                 & \Delta \hat H \le 0 \\
                \exp{\brac{-\beta \Delta \hat H}} & \Delta \hat H > 0
            \end{cases}
            \label{eq:metropolis}
        \end{equation}
        So if a transition lowers the energy it is always done. This
        results in a high ratio between chosen spins and flipped spins.
        Therefore it minimizes the calculations needed for a change of
        the state.

@@

    \subsubsection{Equilibration- and Autocorrelation Time}
    \label{sssec:eqtime}
        To generate an equilibrium state one starts with an arbitrary state
        and waits until it is equilibrated. The count of sweeps till
        equilibrium is called \emph{equilibration time} \(t_{eq}\).
        All measurement should start after this time.\\
        To determine when the system is in equilibrium, one can watch the
        development of some observables and take the point at which there
        are no big changes anymore as equilibrium. In fig.
        \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
        the equilibrium is reached after approximately \(N=50\) sweeps for
        either an initial condition of all spins up and all spins random. It
        does not harm to double that value to be save.
        \begin{figure}[htbp]
            \centering
            \subfigure[Example of an equilibrating Ising system][]{
                    \label{sfig:equiandauto:equiE}
                    \includegraphics[width=0.47\textwidth]{plots/equiE}
            }
            \subfigure[Example of the autocorrelation of an Ising system][]{
                    \label{sfig:equiandauto:autoM}
                    \includegraphics[width=0.47\textwidth]{plots/autoM}
            }
            \caption[Examples for equilibration and autocorrelation]
            {
                \subref{sfig:equiandauto:equiE} Example of a Ising system
                    \(L=64\) reaching equilibrium at \(T=2.36\) and
                \subref{sfig:equiandauto:autoM} the autocorrelation of an
                    Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                    sweeps -- otherwise the decline is too steep to show)
                    on half logarithmic axis.
                    The straight line is an exponential fit \(\exp(-t/\tau)\)
                    with \(\tau = 342(1)\).
            }
            \label{fig:equiandauto}
        \end{figure}\\
        Because every state is generated from the state before, measurements
        of subsequent states are correlated. To determine when two states
        are independent, one calculates the normalized autocorrelation function
        \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
        \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
        logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
        To get the autocorrelation time one can either fit a exponential
        function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
        or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
        \(\tau\) is an estimate after which time two samples are not
        correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
        To make sure that the error is not underestimated one should wait
        \(2\tau\) sweeps between two measurements.
        The autocorrelation time is of course dependent on the temperature.
        For example for the standard Metropolis algorithm the fluctuations
        are strong at high temperatures and subsequent
        states are more dissimilar and therefore less correlated than at low
        temperatures, where less spins are flipping. But the longest
        autocorrelation times are encountered at the critical temperature.
        This effect is called \emph{critical slowing down} and is
        characterized by the \emph{dynamical critical exponent} \(z\)
        \cite{SwendsenWang1987}. The dependence of the autocorrelation time
        \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).
        As mentioned in section \ref{sssec:wolff}, the Wolff cluster algorithm
        decreases \(z\) dramatically. This causes the course of the autocorrelation
        time in dependence on temperature to change significantly as shown in
        section \ref{sec:results}.
